\Title

### 2020-11-18 manuscript/manuscript.tex
Continue edit manuscript

% Define title, authors, affiliations and DOI
% ===========================================
\newcommand{\Title}{
    A better strategy for interpolating gravity and magnetic data
}
\newcommand{\Author}{
    S.R. Soler,
    L. Uieda
}
\newcommand{\AuthorAffil}{
    {\large
        Santiago R. Soler$^{1,2}$,
        Leonardo Uieda$^{3}$
    }
    \\[0.4cm]
    {\small $^{1}$CONICET, Argentina (santiago.r.soler@gmail.com)} \\
    {\small $^{2}$Instituto Geofísico Sismológico Volponi, UNSJ, Argentina} \\
    {\small $^{3}$Department of Earth, Ocean and Ecological Sciences, School of Environmental Sciences, University of Liverpool, UK} \\
}
\newcommand{\DOI}{
    doi:\href{https://doi.org/xxx.xxx/xxxxxx}{xxx.xxx/xxxxxx}
}
\newcommand{\DOILink}{
    \href{https://doi.org/xxx.xxx/xxxxxx}{doi.org/xxx.xxx/xxxxxx}
}

@@

Many variants of the equivalent source technique have been proposed, often
attempting to obtain faster or more accurate solutions.
The key factors that vary between them are: (i) the type of source, (ii)
the location of the sources, and (iii) the solution strategy.
The type of source is most commonly a point mass for gravity or dipole for
magnetics \citep[e.g.,~][]{vonfrese1981, silva1986, mendonca1994, siqueira2017}.
However, right-rectangular prisms \citep[e.g.,][]{barnes2011, jirigalatu2019,
li2020} and tesseroids \citep{bouman2016} have also been used successfully.
In fact, even point sources with a simple inverse distance function, instead of
actual gravity or magnetic fields, can be used as
equivalent sources \citep{cordell1992}.
The sources are often distributed on a regular grid at a constant depth
\citep[e.g.,~][]{leao1989, barnes2011, oliveira2013}
or placed beneath each data point \citep[e.g.,~][]{cordell1992, siqueira2017}.
The model is usually estimated through damped least-squares, which imposes a
heavy computational load when the number of data points is large (e.g.,
airborne and satellite surveys).
To reduce the computational load, \citet{mendonca1994} built the solution
iteratively by incorporating one data point at a time using the ``equivalent
data concept''.
\citet{leao1989} processed the input data using a moving window, only fitting the
data inside the window and predicting observations at the center of the window.
\citet{li2010} and \citet{barnes2011} apply different operations to generate a
sparse representation of the sensitive matrix (respectively, wavelet
compression and quadtree discretization), which significantly improves the
speed of the least-squares solution.
\citet{oliveira2013} parametrized the equivalent layer as a piecewise bivariate
polynomial function, reducing the number of parameters in the solution.
\citet{siqueira2017} developed an iterative solution in which the sensitivity
matrix is transformed into a diagonal matrix with constant terms through the
``excess mass criterion''.
\citet{jirigalatu2019} applied the Gauss-FFT method to speed up the forward
modelling operations and solved the least-squares problem using steepest
descent to avoid calculating the Hessian matrix and solving linear systems.
Most existing methods solve under-determined problems, requiring a much larger
number of equivalent sources than the number of data points.
Furthermore, many of the optimizations proposed are also complex to implement
in a computer program, limiting their wider adoption.

In the present study,
we propose a simple method for reducing the number of equivalent sources.
The survey area is divided into horizontal blocks and one source is assigned
to each block at the median horizontal location of the data points.
For airborne surveys, which are oversampled along tracks, this can greatly
reduce the size of the inverse problem while retaining the quality of
interpolation.
We experiment with the existing strategies of assigning sources at
a uniform depth, a depth relative to the data height, or a depth proportional
to the median distance to adjacent data points.
Through tests on synthetic data, we show that non-uniform depths perform better
at interpolation and that our method is able to achieve the same accuracy as a
traditional equivalent layer while using a fraction of the number of sources.

@@

Reducing the number of sources might help to lower the amount of computer
memory needed to fit their coefficients, although it may be not enough for
a very large set of observation points.
Even if we can greatly reduce the number of sources, having too many data
points would still produce a Jacobian matrix large enough that cannot fit in
memory.
This problem was already faced by other authors (CITE PEOPLE).
Here we propose an iterative algorithm for fitting the source coefficients
using a gradient boosting \citep{friedman2001, friedman2002} strategy.

@@

%Compare the accuracy of the predictions made by each source distribution.
%Need an objective way to compare them.
%Propose a synthetic gravity model made out of prisms of different sizes and
%density contrasts.
%Create a regular grid of observation points where the gravitational effect of
%the synthetic model will be computed.
%Call it the \emph{target grid}.
%Generate synthetic surveys above the same synthetic model and compute the
%gravitational effect of the model on each survey.
%Use each source distribution to predict values based on the ones obtained on
%each survey onto the same grid points of the \emph{target grid}.
%Score each interpolation by comparing the predicted values with the
%\emph{target grid}.
%Finally compare the accuracy of the predictions made by each source
%distribution.

Our first goal is to to compare the performance of all source distributions
from Table~\ref{tab:source-distributions} when predicting harmonic field data
using the equivalent layer technique.
To do so we have created a synthetic model made out of prisms of different
shape and density contrasts.
The gravitational effect of this synthetic model have been computed on
a regular grid (\emph{target} grid) at a constant height and on two different
set of observation points that we will call \emph{synthetic surveys}: a ground
and an airborne survey.
The data from each synthetic survey is interpolated on the points of the
\emph{target} grid and then scored against the \emph{target} grid.
We can then quantitatively compare how accurate is each source distribution at
predicting unobserved data on each synthetic survey and quantify the
computational load needed to do it.

@@

The synthetic model is made out of \NPrisms{} prisms distributed on a region of
\ModelEasting{}$\times$\ModelNorthing{} between \ModelDepth{} deep and the zero
height plane.
Density contrast of prisms range from \ModelMinDensity{} and
\ModelMaxDensity{}.
We created deep big and thin shallow prisms for creating gravity anomalies with
long and short wavelengths, respectively.

The vertical component of the gravity acceleration generated by the synthetic
model has been computed on a regular grid of
\TargetEastingSize{}$\times$\TargetNorthingSize{} points with a spacing of
\TargetSpacing{} located \TargetHeight{} above the zero height plane
(Fig.~\ref{fig:target-grid}).
We will refer to this grid as the \emph{target} grid in the future.

>>>>

% Define title, authors, affiliations and DOI
% ===========================================
\newcommand{\Title}{
    Gradient boosted equivalent sources
}
\newcommand{\Author}{
    S.R. Soler,
    L. Uieda
}
\newcommand{\AuthorAffil}{
    {\large
        Santiago R. Soler$^{1,2}$,
        Leonardo Uieda$^{3}$
    }
    \\[0.4cm]
    {\small $^{1}$CONICET, Argentina (santiago.r.soler@gmail.com)} \\
    {\small $^{2}$Instituto Geofísico Sismológico Volponi, UNSJ, Argentina} \\
    {\small $^{3}$Department of Earth, Ocean and Ecological Sciences, School of Environmental Sciences, University of Liverpool, UK} \\
}
\newcommand{\DOI}{
    doi:\href{https://doi.org/xxx.xxx/xxxxxx}{xxx.xxx/xxxxxx}
}
\newcommand{\DOILink}{
    \href{https://doi.org/xxx.xxx/xxxxxx}{doi.org/xxx.xxx/xxxxxx}
}

@@

Many variants of the equivalent source technique have been proposed, often
attempting to obtain faster or more accurate solutions.
The key factors that vary between them are: (i) the type of source, (ii)
the location of the sources, and (iii) the solution strategy.
The type of source is most commonly a point mass for gravity or dipole for
magnetics \citep[e.g.,~][]{vonfrese1981, silva1986, mendonca1994, siqueira2017}.
However, right-rectangular prisms \citep[e.g.,][]{barnes2011, jirigalatu2019,
li2020} and tesseroids \citep{bouman2016} have also been used successfully.
In fact, even point sources with a simple inverse distance function, instead of
actual gravity or magnetic fields, can be used as
equivalent sources \citep{cordell1992}.
The sources are often distributed on a regular grid at a constant depth
\citep[e.g.,~][]{leao1989, barnes2011, oliveira2013}
or placed beneath each data point \citep[e.g.,~][]{cordell1992, siqueira2017}.
The model is usually estimated through damped least-squares, which imposes a
heavy computational load when the number of data points is large (e.g.,
airborne and satellite surveys).
To reduce the computational load, \citet{mendonca1994} built the solution
iteratively by incorporating one data point at a time using the ``equivalent
data concept''.
\citet{leao1989} processed the input data using a moving window, only fitting the
data inside the window and predicting observations at the center of the window.
\citet{li2010} and \citet{barnes2011} apply different operations to generate a
sparse representation of the sensitive matrix (respectively, wavelet
compression and quadtree discretization), which significantly improves the
speed of the least-squares solution.
\citet{oliveira2013} parametrized the equivalent layer as a piecewise bivariate
polynomial function, reducing the number of parameters in the solution.
\citet{siqueira2017} developed an iterative solution in which the sensitivity
matrix is transformed into a diagonal matrix with constant terms through the
``excess mass criterion''.
\citet{jirigalatu2019} applied the Gauss-FFT method to speed up the forward
modelling operations and solved the least-squares problem using steepest
descent to avoid calculating the Hessian matrix and solving linear systems.
Most existing methods solve under-determined problems, requiring a much larger
number of equivalent sources than the number of data points.
Furthermore, many of the optimizations proposed are also complex to implement
in a computer program, limiting their wider adoption.
In the present study,
we propose a two-way method for reducing the computational load needed to apply
the equivalent sources technique: first by reducing the number of equivalent
sources through a \emph{block-averaging} strategy and then simplifying the
fitting process through a \emph{gradient boosting} algorithm.
The first strategy consists in
dividing the survey area into horizontal blocks and assign one source
to each block at the median horizontal location of the data points.
For airborne surveys, which are oversampled along tracks, this can greatly
reduce the size of the inverse problem while retaining the quality of
interpolation.
The second algorithm simplifies the process of fitting a very large number of
equivalent sources by defining sets of fewer sources, whose coefficients are
fitted iteratively while minimizing a given residue.
Through tests on synthetic data, we show that: (i) the \emph{block-averaged}
sources achieve the same accuracy as a traditional equivalent layer while using
a fraction of the number of sources, and (ii) the \emph{gradient boosting}
algorithm greatly reduces the computational memory required to fit very large
datasets without scarifying predictions' accuracy.
Finally these methods are used to interpolate a 1.7 million gravity data
measurements around Australia onto a regular grid.

@@

Reducing the number of sources might help to lower the amount of computer
memory needed to fit their coefficients, although it may be not enough for
a very large set of observation points.
Even if we can greatly reduce the number of sources, having too many data
points would still produce a Jacobian matrix large enough that cannot fit in
memory.
This problem was already faced by other authors
\citep{
    mendonca1994,
    leao1989,
    li2010,
    barnes2011,
    oliveira2013,
    siqueira2017,
    jirigalatu2019
}.
The iterative method proposed by \citet{siqueira2017} needs equivalent sources
to be located beneath each observation point, so it cannot take advantage of
the block-averaged sources.
\citet{leao1989} perform the fitting simultaneously with the prediction, what
prevents us to use already fitted equivalent sources to make future
predictions. Besides the method is limited to interpolations on regular grids.
\citet{mendonca1994}
\citet{li2010}
\citet{barnes2011}
\citet{jirigalatu2019}

Here we propose an iterative algorithm for fitting the source coefficients
using a gradient boosting \citep{friedman2001, friedman2002} strategy:
(i) define a collection of overlapping windows of equal size that cover the
survey area,
(ii) define one set of equivalent sources per window,
(iii) iteratively fit the coefficients of each set of sources.
Once every set of sources is fitted, we can predict the value of the field on
any observation point as the effect of every set of equivalent sources.

@@

Our first goal is to to compare the performance of all source distributions
from Table~\ref{tab:source-distributions} when predicting harmonic field data
using the equivalent sources technique.
To do so we have created a synthetic model made out of prisms of different
shape and density contrasts.
The gravitational effect of this synthetic model have been computed on
a regular grid (\emph{target} grid) at a constant height and on two different
set of observation points that we will call \emph{synthetic surveys}: a ground
and an airborne survey.
The data from each synthetic survey is interpolated on the points of the
\emph{target} grid and then scored against its values.
We can then quantitatively compare how accurate is each source distribution at
predicting unobserved data on each synthetic survey and quantify the
computational load needed to do it.

@@

The synthetic model is made out of \NPrisms{} prisms distributed on a region of
\ModelEasting{}$\times$\ModelNorthing{} between \ModelDepth{} deep and the zero
height plane.
Density contrast of prisms range from \ModelMinDensity{} and
\ModelMaxDensity{}.
We built deep big and thin shallow prisms for creating gravity anomalies with
long and short wavelengths, respectively.

The vertical component of the gravity acceleration generated by the synthetic
model has been computed on a regular grid of
\TargetEastingSize{}$\times$\TargetNorthingSize{} points with a spacing of
\TargetSpacing{} located \TargetHeight{} above the zero height plane
(Fig.~\ref{fig:target-grid}), using \citet{nagy2000} and \citet{nagy2002}.
We will refer to this grid as the \emph{target} grid in the future.
